---
title: "EDS 216: Meta-Analysis and Systematic Reviews"
categories:
  - course reflections
description: |
  Data aggregation and reviews and assessments of existing scientific literature!
author:
  - name: Scout Leonard
    url: {https://scoutcleonard.github.io}
date: 10-22-2021
output:
  distill::distill_article:
    self_contained: false
preview: files/systematic_review.jpg
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

My final course of the summer quarter was EDS 216: Meta-Analysis and Systematic Review with Professors Scott Jasechko. In the course, we began with studying existing systematic reviews of environmental data science topics. We presented one minute presentations about systematic reviews we read; a true test of understanding and fast-talking! :)

Each of my classmates were asked to develop a question with the structure, "What is the effect of X on Y?" These investigations were to be used to develop our own systematic review protocols for the duration of the class. 

The question I chose was based on my previous work in urban food systems in Oakland Unified School District's Nutrition Services Program. I asked, "What are the effects of Farm to School programs on student well-being. My systematic review protocol is in the PDF below. 

```{r, fig.width = 7, fig.height = 10}
knitr::include_graphics("files/eds216_SystematicReview_ScoutLeonard.pdf")
```

One of the main skills I learned in this class was a protocol for conducting literature searches by creating search strings from defined PICO terms. PICO terms represent the Population, Intervention, Comparative, and Outcome of your question of interest. Using Boolean operators and wildcards, you can conduct searches for literature that meets the structure of your scientific question. 

Systematic reviews are vitally important. Science is expensive, and knowing if and how your question has been answered can prevent duplicate research or inform better, further exploration. Also, scientific literature is massively complex; systematic reviews can be helpful briefs for policy-makers to understand a body of literature without having to read it all themselves!

Distill is a publication format for scientific and technical writing, native to the web.

Learn more about using Distill at <https://rstudio.github.io/distill>.


