---
title: "EDS 221: Things I'm Learning!"
description: |
  A reflection on all the things I'm learning in my first weeks of graduate school.
author:
  - name: Scout Leonard
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(palmerpenguins)
```

## **I'm learning lots in Environmental Data Science 221: Scientific Programming Essentials!**

Here are some examples of things I've learned so far: 

One **function** we've learned that I enjoy is the `case_when()` function. When paired with the `mutate()` function, we can sort data in a newly created column based on existing data. Below is an example with the Palmer Penguins dataset. 

```{r}
case_example <- penguins %>% 
  mutate(binned_flipper_size = case_when(
    flipper_length_mm < 185 ~ "small", 
    flipper_length_mm >= 185 ~ "medium"
  ))


ggplot(data = case_example, aes(x = species, y = flipper_length_mm, fill = binned_flipper_size)) +
  geom_col(position = "fill") +
  labs(title = "Playing with case_when(), a graphic example!")
```
I made this stacked bar graph so you can see that I have a new column that has two classifications for flipper length: *medium* and *small*.

A **concept** we've learned in EDS 221 is tidy data. As an organizational fiend, this has been one of my favorite concepts! I have worked on many, many untidy spreadsheets over the years. In tidy data, data is organized in a predictable way! Most importantly, tidy data has: 

1.) Each variable is a column.
2.) Each observation is a row. 
3.) Each cell contains **1** value. 



Distill is a publication format for scientific and technical writing, native to the web.

Learn more about using Distill at <https://rstudio.github.io/distill>.


